---
title: "Data Normalization for Multi-Omics"
output:
  html_document:
    df_print: paged
---

# Data Normalization for Multi-Omics

**Learning Objectives**

By the end of this 30-minute exercise you will be able to:

1.  Explain *why* normalization is necessary before comparing samples in omics experiments.
2.  Compute **RPKM** and **TPM** for RNA-seq read counts and describe when each is appropriate.
3.  Apply **quantile normalization** to intensity-based data (microarray / proteomics).
4.  Apply **Pareto scaling** to metabolomics data and contrast it with autoscaling.

**Data file:** `gene_treatment_data.txt` (5 genes, 6 samples, with gene lengths)

**Structure:** Each section has a worked example followed by a short exercise. Exercises are "fill in the blank" — look for lines marked `### YOUR CODE HERE ###`.

```{r setup, message=FALSE}
library(tidyverse)
library(reshape2)

# Consistent plot style for the session
theme_set(theme_bw(base_size = 13))
```

------------------------------------------------------------------------

## Section 1: RNA-seq Normalization — RPKM and TPM (\~10 min)

### Why normalize RNA-seq counts?

Raw read counts for a gene depend on two **technical factors** that have nothing to do with biology:

| Bias | Cause | Fix |
|------------------------|----------------------------|--------------------|
| **Sequencing depth** | A sample sequenced more deeply produces more reads for *every* gene. | Divide by total reads (per-million scaling). |
| **Gene length** | Longer transcripts capture more reads even at equal expression. | Divide by gene length (per-kilobase scaling). |

**RPKM** (Reads Per Kilobase per Million mapped reads) corrects for both:

$$\text{RPKM} = \frac{\text{read count}}{\left(\frac{\text{gene length}}{10^3}\right) \times \left(\frac{\text{total reads}}{10^6}\right)}$$

**TPM** (Transcripts Per Million) also corrects for both biases, but normalizes *per sample* so that the column totals are always 1,000,000, making samples directly comparable:

$$\text{TPM}_i = \frac{\text{read count}_i \;/\; \text{gene length}_i}{\sum_j \left(\text{read count}_j \;/\; \text{gene length}_j\right)} \times 10^6$$

> **Rule of thumb:** TPM is preferred in modern analyses because RPKM column sums differ between samples, which can distort cross-sample comparisons.

```{r load-data}
# Load the gene expression data (5 genes, 6 samples + gene length column)
gene_data <- read.delim("gene_treatment_data.txt", row.names = 1)
cat("Shape:", nrow(gene_data), "x", ncol(gene_data), "\n")
gene_data
```

```{r raw-plots, fig.width=12, fig.height=3.5}
# Separate the count columns from the gene length column
counts <- gene_data[, 1:6]   # 5 genes x 6 samples
lengths <- gene_data$length   # gene lengths in base pairs

# Visualize the raw data — notice how samples have different totals (sequencing depth)
par(mfrow = c(1, 2))

# Bar plot of raw read counts per sample
counts_long <- melt(as.matrix(counts))
colnames(counts_long) <- c("gene", "sample", "reads")
p1 <- ggplot(counts_long, aes(x = sample, y = reads, fill = gene)) +
  geom_col(position = "dodge") +
  labs(title = "Raw Read Counts per Sample", y = "Reads") +
  theme(axis.text.x = element_text(angle = 0))

# Total reads per sample
depth <- data.frame(sample = colnames(counts), total = colSums(counts))
p2 <- ggplot(depth, aes(x = sample, y = total)) +
  geom_col(fill = "steelblue") +
  labs(title = "Total Reads per Sample (Sequencing Depth)", y = "Total Reads")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

```{r rpkm}
# --- RPKM calculation (vectorized) ---
# total_reads: sum of counts per sample (column sums)
total_reads <- colSums(counts)

# Divide each count by (gene_length_in_kb * total_reads_in_millions)
rpkm <- sweep(counts, 1, lengths / 1e3, "/")
rpkm <- sweep(rpkm, 2, total_reads / 1e6, "/")

cat("RPKM values:\n")
round(rpkm, 1)
```

```{r tpm}
# --- TPM calculation ---
# Step 1: compute the rate = reads per kilobase of gene
rate <- sweep(counts, 1, lengths / 1e3, "/")

# Step 2: scale each sample (column) so that the sum equals 1,000,000
tpm <- sweep(rate, 2, colSums(rate), "/") * 1e6

cat("TPM column sums (should each be 1,000,000):\n")
print(round(colSums(tpm), 1))
cat("\n")
round(tpm, 1)
```

```{r rpkm-vs-tpm, fig.width=12, fig.height=3.5}
# Compare RPKM vs TPM side by side
rpkm_long <- melt(as.matrix(rpkm))
colnames(rpkm_long) <- c("gene", "sample", "value")
p1 <- ggplot(rpkm_long, aes(x = sample, y = value, fill = gene)) +
  geom_col(position = "dodge") +
  labs(title = "After RPKM Normalization", y = "RPKM") +
  theme(axis.text.x = element_text(angle = 0))

tpm_long <- melt(as.matrix(tpm))
colnames(tpm_long) <- c("gene", "sample", "value")
p2 <- ggplot(tpm_long, aes(x = sample, y = value, fill = gene)) +
  geom_col(position = "dodge") +
  labs(title = "After TPM Normalization", y = "TPM") +
  theme(axis.text.x = element_text(angle = 0))

gridExtra::grid.arrange(p1, p2, ncol = 2)

# Notice: RPKM column sums differ, TPM column sums are all 1,000,000
cat("RPKM column sums:", round(colSums(rpkm), 1), "\n")
cat("TPM  column sums:", round(colSums(tpm), 1), "\n")
```

### Exercise 1 — Compute RPKM and TPM yourself

The cell below creates a small synthetic dataset with 3 genes and 2 samples. Fill in the missing lines to compute RPKM and TPM.

```{r exercise-1}
# Synthetic data
ex_counts <- data.frame(
  sample_A = c(200, 100, 50),
  sample_B = c(300, 80, 70),
  row.names = c("geneX", "geneY", "geneZ")
)
ex_lengths <- c(geneX = 2000, geneY = 500, geneZ = 1000)

cat("Counts:\n")
print(ex_counts)
cat("\nGene lengths:\n")
print(ex_lengths)

# --- RPKM ---
ex_total <- colSums(ex_counts)
# ex_rpkm <- ### YOUR CODE HERE ###   # one-line vectorized RPKM formula

# cat("\nRPKM:\n")
# print(round(ex_rpkm, 1))

# --- TPM ---
# ex_rate <- ### YOUR CODE HERE ###    # reads divided by gene length in kb
# ex_tpm  <- ### YOUR CODE HERE ###    # scale so each column sums to 1e6

# cat("\nTPM:\n")
# print(round(ex_tpm, 1))
# cat("TPM column sums:", round(colSums(ex_tpm), 1), "\n")
```

------------------------------------------------------------------------

## Section 2: Quantile Normalization — Microarray / Proteomics (\~8 min)

### When and why

Microarray and label-free proteomics produce **intensity** values. Technical variation (dye bias, detector differences) causes the intensity *distributions* to differ across samples even when the underlying biology is the same.

**Quantile normalization** forces every sample to have the *same* distribution by:

1.  **Sort** each column independently.
2.  **Average** across each row of the sorted matrix.
3.  **Replace** each original value with the average that matches its rank.

> This is a strong assumption — it works well when most features are *not* differentially expressed or abundant.

```{r qn-toy}
# --- Small worked example (4 genes, 3 samples) ---
df_toy <- data.frame(
  C1 = c(5, 2, 3, 4),
  C2 = c(4, 1, 4, 2),
  C3 = c(3, 4, 6, 8),
  row.names = c("A", "B", "C", "D")
)
cat("Original data:\n")
df_toy
```

```{r qn-steps}
# Step 1: Sort each column independently
df_sorted <- as.data.frame(apply(df_toy, 2, sort))
rownames(df_sorted) <- rownames(df_toy)
cat("Step 1 — Sorted columns:\n")
df_sorted

# Step 2: Compute the row means of the sorted data
row_means <- rowMeans(df_sorted)
names(row_means) <- 1:length(row_means)  # label by rank
cat("\nStep 2 — Row means (one value per rank):\n")
print(row_means)

# Step 3: Map each original value's rank to the corresponding row mean
df_qn <- as.data.frame(apply(df_toy, 2, function(col) {
  ranks <- rank(col, ties.method = "min")
  row_means[ranks]
}))
rownames(df_qn) <- rownames(df_toy)
cat("\nStep 3 — Quantile-normalized data:\n")
df_qn
```

```{r qn-function}
# Wrap it into a reusable function
quantile_normalize <- function(df) {
  # Quantile-normalize columns of a numeric data frame.
  df_sorted <- as.data.frame(apply(df, 2, sort))
  row_means <- rowMeans(df_sorted)
  names(row_means) <- 1:length(row_means)
  df_qn <- as.data.frame(apply(df, 2, function(col) {
    ranks <- rank(col, ties.method = "min")
    row_means[ranks]
  }))
  rownames(df_qn) <- rownames(df)
  return(df_qn)
}
```

```{r qn-proteomics, fig.width=12, fig.height=3.5}
# Test on simulated proteomics data: 3 samples with different Poisson means
set.seed(42)
df_prot <- data.frame(
  sample_1 = rpois(5000, lambda = 10),
  sample_2 = rpois(5000, lambda = 15),
  sample_3 = rpois(5000, lambda = 20)
)

df_prot_qn <- quantile_normalize(df_prot)

# Before / after density plots
prot_long <- melt(as.matrix(df_prot))
colnames(prot_long) <- c("feature", "sample", "value")
p1 <- ggplot(prot_long, aes(x = value, color = sample)) +
  geom_density(linewidth = 0.8) +
  labs(title = "Before Quantile Normalization")

prot_qn_long <- melt(as.matrix(df_prot_qn))
colnames(prot_qn_long) <- c("feature", "sample", "value")
p2 <- ggplot(prot_qn_long, aes(x = value, color = sample)) +
  geom_density(linewidth = 0.8) +
  labs(title = "After Quantile Normalization")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

### Exercise 2 — Quantile-normalize a new dataset

Generate 4 samples of 2000 features each, drawn from Poisson distributions with means 5, 12, 25, and 40. Apply quantile normalization, then make a before/after density plot.

```{r exercise-2, fig.width=12, fig.height=3.5}
set.seed(7)

# df_ex2 <- data.frame(
#   S1 = rpois(2000, lambda = 5),
#   S2 = rpois(2000, lambda = 12),
#   S3 = ### YOUR CODE HERE ###,
#   S4 = ### YOUR CODE HERE ###
#)

# df_ex2_qn <- ### YOUR CODE HERE ###   # apply quantile normalization

# Plot before and after
### YOUR CODE HERE ###   # density plot before (left) and after (right)
```

------------------------------------------------------------------------

## Section 3: Scaling for Metabolomics — Pareto and Autoscaling (\~8 min)

### The problem

LC-MS metabolomics data spans several orders of magnitude. A few high-abundance metabolites dominate the variance, masking biologically interesting low-abundance signals.

**Scaling** (applied *per feature*, across samples) addresses this:

| Method | Formula (for feature *i*) | Effect |
|----------------|-----------------------------------------|----------------|
| **Autoscaling** (z-score) | $(x - \bar{x}) / s$ | Every feature gets unit variance. Can inflate noisy features. |
| **Pareto scaling** | $(x - \bar{x}) / \sqrt{s}$ | Down-weights large features less aggressively than autoscaling. Preferred for metabolomics. |

> **Note:** Scaling operates across *samples for each feature* (row-wise when features are rows), unlike quantile normalization which operates on whole columns.

```{r scaling, fig.width=16, fig.height=3.5}
# Simulate 6 metabolomics samples with power-law-like distributions
# Using beta distribution to mimic scipy.stats.powerlaw(a) which has CDF x^a on [0,1]
set.seed(21)
power_rvs <- function(a, n) {
  # power-law distribution on [0,1]: CDF = x^a, inverse CDF = u^(1/a)
  runif(n)^(1 / a)
}

df_met <- data.frame(
  sample_1 = power_rvs(0.5, 5000),
  sample_2 = power_rvs(0.7, 5000),
  sample_3 = power_rvs(1.5, 5000),
  sample_4 = power_rvs(2.0, 5000),
  sample_5 = power_rvs(1.3, 5000),
  sample_6 = power_rvs(0.3, 5000)
)

# Feature-wise statistics (each row = one metabolite measured across 6 samples)
row_mean <- rowMeans(df_met)
row_std  <- apply(df_met, 1, sd)

# Pareto scaling: subtract mean, divide by sqrt(std)
df_pareto <- sweep(df_met, 1, row_mean, "-")
df_pareto <- sweep(df_pareto, 1, sqrt(row_std), "/")

# Autoscaling (z-score): subtract mean, divide by std
df_auto <- sweep(df_met, 1, row_mean, "-")
df_auto <- sweep(df_auto, 1, row_std, "/")

# Compare all three
met_long <- melt(as.matrix(df_met))
colnames(met_long) <- c("feature", "sample", "value")
p1 <- ggplot(met_long, aes(x = value, color = sample)) +
  geom_density(linewidth = 0.8) +
  labs(title = "Before Scaling") +
  theme(legend.position = "none")

par_long <- melt(as.matrix(df_pareto))
colnames(par_long) <- c("feature", "sample", "value")
p2 <- ggplot(par_long, aes(x = value, color = sample)) +
  geom_density(linewidth = 0.8) +
  labs(title = "After Pareto Scaling") +
  theme(legend.position = "none")

auto_long <- melt(as.matrix(df_auto))
colnames(auto_long) <- c("feature", "sample", "value")
p3 <- ggplot(auto_long, aes(x = value, color = sample)) +
  geom_density(linewidth = 0.8) +
  labs(title = "After Autoscaling (z-score)") +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

### Exercise 3 — Implement Pareto scaling

Using the `df_met` data frame from above, implement Pareto scaling yourself. Then answer the question at the bottom.

```{r exercise-3, fig.width=10, fig.height=3.5}
# Compute the row (feature) mean and standard deviation
# feat_mean <- ### YOUR CODE HERE ###
# feat_std  <- ### YOUR CODE HERE ###

# Pareto scaling: subtract mean, divide by sqrt of std
# df_pareto_ex <- ### YOUR CODE HERE ###

# Quick check — plot one sample before and after
# par_before <- data.frame(value = df_met$sample_1)
# par_after  <- data.frame(value = df_pareto_ex$sample_1)

# p1 <- ggplot(par_before, aes(x = value)) +
#   geom_density(linewidth = 0.8, color = "coral") +
#   labs(title = "sample_1 — Raw")
# 
# p2 <- ggplot(par_after, aes(x = value)) +
#   geom_density(linewidth = 0.8, color = "teal") +
#   labs(title = "sample_1 — Pareto Scaled")
# 
# gridExtra::grid.arrange(p1, p2, ncol = 2)

# QUESTION: Why might you prefer Pareto scaling over autoscaling
# for metabolomics data?
#
### YOUR ANSWER HERE ###
```

------------------------------------------------------------------------

## Section 4: Summary and Reflection (\~4 min)

| Omics type | Typical normalization | Key idea |
|-------------------|-----------------------------------|------------------|
| RNA-seq | RPKM / TPM | Correct for **gene length** and **sequencing depth** |
| Microarray / Proteomics | Quantile normalization | Force all samples to share the **same distribution** |
| Metabolomics | Pareto / Autoscaling | Reduce dominance of **high-variance features** |

### Key take-aways

-   **Always visualize** your data before and after normalization.
-   No single method works for all data types — the right choice depends on the technology and its specific biases.
-   TPM is generally preferred over RPKM for RNA-seq because column totals are consistent across samples.

### Exercise 4 — Match the method

Replace each `"???"` with the correct normalization method name from: `"RPKM"`, `"TPM"`, `"Quantile"`, `"Pareto"`, `"Autoscaling"`

```{r exercise-4}
answers <- c(
  "Corrects for gene length and sequencing depth; column sums vary"          = "???",
  "Corrects for gene length and sequencing depth; column sums are 1e6"       = "???",
  "Forces identical distributions across samples"                            = "???",
  "Mean-centers and divides by sqrt(sd); good for metabolomics"              = "???",
  "Mean-centers and divides by sd; can inflate noisy low-abundance features" = "???"
)

for (i in seq_along(answers)) {
  cat(sprintf("  %12s  <-  %s\n", answers[i], names(answers)[i]))
}
```

------------------------------------------------------------------------

### Further Reading

-   Zhao et al. (2020) — [TPM, FPKM, or Normalized Counts? A Comparative Study](https://pmc.ncbi.nlm.nih.gov/articles/PMC8220791/)
-   Hicks & Irizarry (2015) — [Quantile normalization assumptions and consequences](https://www.biorxiv.org/content/10.1101/012203v1.full)
-   van den Berg et al. (2006) — [Centering, scaling, and transformations for metabolomics](https://pmc.ncbi.nlm.nih.gov/articles/PMC1534033/)

```{r done}
cat("DONE\n")
```
